A self-consistency check for unitary propagation of Hawking quanta

### 2016-08-16 Draft_BH_shells.tex

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}
\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicitly conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschid geometry given by general relativity is a valid description to the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Local quantum field theory:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests as a unitary propagation of every near-horizon mode into every asymptotic Hawking quantum.
\end{itemize}

@@

Fortunately, Feynman, in his famous lectures, presented an interesting trick to circumvent this obstacle. 
This trick, to a certain extend, enables us to perform a {\bf self-consistency check} of the classical-background assumption {\bf without} knowing the quantum nature of the background.
In Sec.\ref{sec-DoubleSlit}, we will review Feynman's analysis of the double-slit experiment. 
In his case, the double-slit is the classical background, the particle passing through it is the quantum subsystem, and subsystem unitarity is checked by whether an interference pattern can be observed on the final projection screen. 
He showed that we can ignore the quantum details of the double-slit and summarize that as an uncertainty in its position, which is a classical quantity. 
When such uncertainty is fixed, a classical calculation can show that the interference pattern is always visible as long as we put the final projection screen to be infinitely far away. 
It is in this sense that a unitary evolution of the subsystem is an arbitrarily good approximation.

@@

\begin{itemize}
\item {\bf Classical coherence by tracking geodesics.} The classical mode of a massless field basically has its peaks and nodes following null geodesics.
Thus, instead of actually solving the mode function, we can perform a much simpler calculation to track geodesics.
We start with two out-going null geodesics with the separation of one period of such mode near the horizon, and then we calculate the change in their separations at spatial infinity.
If this change is comparable to the wavelength, then the mode function of this wavelength loses classical coherence.
\item {\bf The background and the uncertainty.} The classical background in this case is the Schwarzschild geometry, which can be written as the metric $g_{\mu\nu}$. 
A classical uncertainty can be expressed as deviation from such metric, $\Delta g_{\mu\nu}$. 
There are two major challenges here.
\begin{enumerate}
\item There are many more variables than the double-slit experiment. 
We consider only one particular form of $\Delta g_{\mu\nu}$ in this paper---back-reactions from the presence of extra matter of zero total energy. 
It is simple to calculate and reflects the physcial intuition of vacuum pair fluctuation.
\item The value of $\Delta g_{\mu\nu}$ is gauge dependent. 
By relating it to the presence of extra matter, we can express it to a gauge-invariant local quantity, which is the natural thing to hold fixed in our analysis.
\end{enumerate}
Note that if we were reporting a positive result, that such uncertainty upholds unitarity, then one should question the validaty of learning a general lesson from a special example. However, we will show that this particular form of $\Delta g_{\mu\nu}$ is already sufficient to question unitarity, and we did not choose this form to specifically do that. Thus our limited analysis is sufficient to raise a reasonable doubt to the classical-background approximation.
\item {\bf Renormalization.} It is well-known that na\"ive applications of QFT often suffer from UV/IR divergences.
In our case, we will see that even in Minkowski space, geometric uncertainty already leads to a finite change between the two null geodesics.
We will simply ``renormalize'' that away by saying that if the black hole case leads to a similar change, it should not be taken as a serious problem.
Our surprising result is that in the black hole case, we get an infinitely larger change, which is difficult to blame on renormalization.
\end{itemize}

@@

As far as we know, an observable that explicitly checks the subsystem unitarity of local QFT is the cosmological Bell inequality \cite{Mal15}. 
One can apply our self-consistency check to such model and update the prediction of whether a violation of Bell inequality is actually expected.

@@

{\bf XXX Modify the content accordingly to fit their status in the appendix: Darsh/Daniel.}

@@

Since $\delta t_2$ is equal to the time interval observed by the observer at $r_0$, $\delta t_0$, we can find the proper time observed by the observer, $\delta \tau_0$,

\begin{equation}
	\delta \tau_0  = \delta t_e \left( \left( \frac{r_2 - 2M}{r_1 - 2M} \right) \left( \frac{r_1 - 2(M-\Delta M)}{r_2 - 2(M - \Delta M)} \right) \left( 1 -\frac{2M}{r_0} \right) \right)^\frac{1}{2}. 
\end{equation}

Since we assume the observer is far away we can neglect the term $\frac{2M}{r_0}$. The difference between the proper time interval in the presence of shells to the value without shells is

\begin{equation}
	\delta \tau_0 - \delta \bar{\tau_0} = \delta t_e \left( \left( \frac{r_2 - 2M}{r_1 - 2M} \right)^\frac{1}{2} \left( \frac{r_1 - 2(M -\Delta M)}{r_2 - 2(M- \Delta M)} \right)^\frac{1}{2} - 1\right),
\end{equation}

since $\delta t_e = \delta \bar{\tau_0}$ when the observer is far away, we can define a dimensionless quantity,

@@

where $K_{ab}$ is the extrinsic curvature corresponding to the induced metric we are using, $K = K_{ab} h^{ab}$ is the trace of the extrinsic curvature taken with respect to the induced metric and the notation of square brackets represents the difference in a quantity in two different induced metrics. We can define the induced metrics for constant $r$ values; for region $I$ it is

\begin{equation}
	ds_{3M}^2 = - \left( 1 - \frac{2M}{r} \right)^{-1} dt^2 + r^2 d \Omega_2^2.	\label{13}
\end{equation}

Analogously we have the metric corresponding to mass $M' = M- \Delta M$ in region $II$

\begin{equation}
	ds_{3M'}^2 = - \left( 1 - \frac{2M'}{r} \right)^{-1} dt'^2 + r^2 d \Omega_2^2.	\label{14}
\end{equation}

We could rewrite Eq (\ref{13}) and (\ref{14}) in Gaussian normal coordinates (where the coefficient in front of the time component is 1), however the first order junction condition is that the induced metric on both sides of the shell must be the same. Therefore it is easy to find a relation between the two time coordinates

\begin{equation}
	dt'^2 = \left( \frac{1 - \frac{2M'}{r_1}}{ 1- \frac{2M}{r_1}} \right) dt^2.
\end{equation}

Using this, we calculate the extrinsic curvature components

\begin{eqnarray}
	(K^t_t)^{(M)} & = & \frac{M}{r_1^2} \left( 1 - \frac{2M}{r_1} \right)^{-\frac{1}{2}} 	\nonumber	\\
	(K^\theta_\theta)^{(M)} & = & \frac{1}{r_1} \left( 1 - \frac{2M}{r_1} \right)^\frac{1}{2} 
	= (K^\phi_\phi)^{(M)}	\label{16}
\end{eqnarray}

where the superscript of $M$ denotes the quantities evaluated in the metric of mass $M$. Analogous expressions hold for the extrinsic curvature components calculated in the metric of mass $M'$ by just replacing $M$ by $M'$ in Eq (\ref{16}). With these ingredients we compute the $00$ component of the surface stress energy tensor

@@

Eqs (\ref{sig}, \ref{20}, \ref{21}) can be used in Eq (\ref{12}) to replace $\Delta M, \epsilon, \delta \epsilon$ respectively by the appropriate physical quantities. It is not possible to do this analytically but it can be done numerically. We show the plots of $\frac{\delta \tau_0 - \delta \bar{\tau_0}}{\delta \bar{\tau_0}}$ as functions of $M$. 
\\
\\
We can obtain analytic expressions in the limit that the shells are very close to the horizon and this is what is presented in the next section.

@@

\begin{figure}[h!]
\begin{center}.
\includegraphics[scale = 0.5]{tau_compare.pdf}
\caption{}
\label{}
\end{center}
\end{figure}

>>>>

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}
\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicitly conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschid geometry given by general relativity is a valid description to the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Local quantum field theory:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests as a unitary propagation of every near-horizon mode into every asymptotic Hawking quanta.
\end{itemize}

@@

Fortunately, Feynman, in his famous lectures, presented an interesting trick to circumvent this obstacle. 
This trick, to a certain extent, enables us to perform a {\bf self-consistency check} of the classical-background assumption {\bf without} knowing the quantum nature of the background.
In Sec.\ref{sec-DoubleSlit}, we will review Feynman's analysis of the double-slit experiment. 
In his case, the double-slit is the classical background, the particle passing through it is the quantum subsystem, and subsystem unitarity is checked by whether an interference pattern can be observed on the final projection screen. 
He showed that we can ignore the quantum details of the double-slit and summarize that as an uncertainty in its position, which is a classical quantity. 
When such uncertainty is fixed, a classical calculation can show that the interference pattern is always visible as long as we put the final projection screen to be infinitely far away. 
It is in this sense that a unitary evolution of the subsystem is an arbitrarily good approximation.

@@

\begin{itemize}
\item {\bf Classical coherence by tracking geodesics.} The classical mode of a massless field basically has its peaks and nodes following null geodesics.
Thus, instead of actually solving the mode function, we can perform a much simpler calculation to track geodesics.
We start with two out-going null geodesics with the separation of one period of such mode near the horizon, and then we calculate the change in their separations at spatial infinity.
If this change is comparable to the wavelength, then the mode function of this wavelength loses classical coherence.
\item {\bf The background and the uncertainty.} The classical background in this case is the Schwarzschild geometry, which can be written as the metric $g_{\mu\nu}$. 
A classical uncertainty can be expressed as deviation from such metric, $\Delta g_{\mu\nu}$. 
There are two major challenges here.
\begin{enumerate}
\item There are many more variables than the double-slit experiment. 
We consider only one particular form of $\Delta g_{\mu\nu}$ in this paper---back-reactions from the presence of extra matter of zero total energy. 
It is simple to calculate and reflects the physcial intuition of vacuum pair fluctuation.
\item The value of $\Delta g_{\mu\nu}$ is gauge dependent. 
By relating it to the presence of extra matter, we can express it as a gauge-invariant local quantity, which is the natural thing to hold fixed in our analysis.
\end{enumerate}
Note that if we were reporting a positive result, that such uncertainty upholds unitarity, then one should question the validaty of learning a general lesson from a special example. However, we will show that this particular form of $\Delta g_{\mu\nu}$ is already sufficient to question unitarity, and we did not choose this form to specifically do that. Thus our limited analysis is sufficient to raise a reasonable doubt to the classical-background approximation.
\item {\bf Renormalization.} It is well-known that na\"ive applications of QFT often suffer from UV/IR divergences.
In our case, we will see that even in Minkowski space, geometric uncertainty already leads to a finite change between the two null geodesics.
We will simply ``renormalize'' that away by saying that if the black hole case leads to a similar change, it should not be taken as a serious problem.
Our surprising result is that in the black hole case, we get an infinitely larger change, which is difficult to blame on renormalization.
\end{itemize}

@@

As far as we know, an observable that explicitly checks the subsystem unitarity of local QFT is the cosmological Bell inequality \cite{Mal15}. 
One can apply our self-consistency check to such a model and update the prediction of whether a violation of Bell inequality is actually expected.

@@

@@

Since $\delta t_2$ is equal to the time interval observed by the observer at $r_0$, $\delta t_0$, we can find the proper time observed by the observer, $\delta \tau_0  = \delta t_e \left( \left( \frac{r_2 - 2M}{r_1 - 2M} \right) \left( \frac{r_1 - 2(M-\Delta M)}{r_2 - 2(M - \Delta M)} \right) \right)^\frac{1}{2}$. Where we have dropped the $\left( 1 - \frac{2M}{r_0} \right)^\frac{1}{2}$ term since we assume the observer is very far away. $\delta t_e = \delta \bar{\tau_0}$ when the observer is far away therefore we can define a dimensionless quantity that quantifies the change in proper time intervals for consecutive photons in perturbed spacetime case and the unperturbed case,

@@

where $K_{ab}$ is the extrinsic curvature corresponding to the induced metric we are using, $K = K_{ab} h^{ab}$ is the trace of the extrinsic curvature taken with respect to the induced metric and the notation of square brackets represents the difference in a quantity in two different induced metrics. 
%
%We can define the induced metrics for constant $r$ values; for region $I$ it is 
%
%\begin{equation}
%	ds_{3M}^2 = - \left( 1 - \frac{2M}{r} \right)^{-1} dt^2 + r^2 d \Omega_2^2.	\label{13}
%\end{equation}
%
%Analogously we have the metric corresponding to mass $M' = M- \Delta M$ in region $II$
%
%\begin{equation}
%	ds_{3M'}^2 = - \left( 1 - \frac{2M'}{r} \right)^{-1} dt'^2 + r^2 d \Omega_2^2.	\label{14}
%\end{equation}
%
%We could rewrite Eq (\ref{13}) and (\ref{14}) in Gaussian normal coordinates (where the coefficient in front of the time component is 1), however the first order junction condition is that the induced metric on both sides of the shell must be the same. Therefore it is easy to find a relation between the two time coordinates
%
%\begin{equation}
%	dt'^2 = \left( \frac{1 - \frac{2M'}{r_1}}{ 1- \frac{2M}{r_1}} \right) dt^2.
%\end{equation}
%
%Using this, we calculate the extrinsic curvature components
%
%\begin{eqnarray}
%	(K^t_t)^{(M)} & = & \frac{M}{r_1^2} \left( 1 - \frac{2M}{r_1} \right)^{-\frac{1}{2}} 	\nonumber	\\
%	(K^\theta_\theta)^{(M)} & = & \frac{1}{r_1} \left( 1 - \frac{2M}{r_1} \right)^\frac{1}{2} 
%	= (K^\phi_\phi)^{(M)}	\label{16}
%\end{eqnarray}
%
%where the superscript of $M$ denotes the quantities evaluated in the metric of mass $M$. Analogous expressions hold for the extrinsic curvature components calculated in the metric of mass $M'$ by just replacing $M$ by $M'$ in Eq (\ref{16}).

The $tt$ component of the surface stress energy tensor is

@@

Eqs (\ref{sig}, \ref{20}, \ref{21}) can be used in Eq (\ref{12}) to replace $\Delta M, \epsilon, \delta \epsilon$ respectively by the appropriate physical quantities. It is not possible to do this analytically but it can be done numerically. We show the plots of $\frac{\delta \tau_0 - \delta \bar{\tau_0}}{\delta \bar{\tau_0}}$ as functions of $M$. We can obtain analytic expressions for Eq (\ref{12}) in the limit that the shells are very close to the horizon and this is what is presented in the next section.

@@
